---
title: "02. R Setup for Genomics"
author: "Muhammad Aziz Muslim"
date: "last-modified"
categories: [setup, r-programming]
---

## Learning Objectives

By the end of this tutorial, you will:

- Have a complete R/RStudio setup for genomic analysis
- Know essential R packages for breeding genomics
- Understand R project organization
- Be able to create reproducible analyses

## Installing R and RStudio

### Step 1: Install R

Download from: https://cran.r-project.org/

**Windows:**
1. Download `.exe` installer
2. Run and follow installation wizard
3. Install to default location

**macOS:**
```bash
# Using Homebrew (recommended)
brew install r

# Or download from CRAN
# https://cran.r-project.org/bin/macosx/
```

**Linux (Ubuntu/Debian):**
```bash
sudo apt-get update
sudo apt-get install r-base r-base-dev
```

### :::{.callout-tip}
**Pro Tip:** Always use the latest version of R (currently 4.3+) for best package compatibility.
:::

### Step 2: Install RStudio

Download from: https://posit.co/download/rstudio-desktop/

RStudio provides:
- Better code editor
- Integrated help
- Project management
- Git integration
- Package management

## Essential Package Installation

Create a setup script for reproducibility:

```r
# =============================================================================
# File: 00_setup.R
# Description: Install and load all required packages for genomics
# =============================================================================

# Function to check and install packages
install_if_missing <- function(packages) {
  new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
  
  if(length(new_packages)) {
    message(paste("Installing:", paste(new_packages, collapse = ", ")))
    install.packages(new_packages, dependencies = TRUE)
  }
}

# Core tidyverse packages
core_packages <- c(
  "tidyverse",    # Data manipulation and visualization
  "data.table",   # Fast data processing
  "readxl",       # Excel file reading
  "writexl",      # Excel file writing
  "haven"         # SPSS/Stata/SAS files
)

install_if_missing(core_packages)

# Genomics-specific packages
genomics_packages <- c(
  "rrBLUP",       # Ridge regression BLUP
  "BGLR",         # Bayesian regression
  "sommer",       # Mixed models
  "qtl",          # QTL mapping
  "qtl2",         # Next-gen QTL mapping
  "gap",          # Genetic analysis
  "genio",        # PLINK file I/O
  "snpStats",     # SNP data handling
  "qqman"         # Manhattan and QQ plots
)

# Install genomics packages
install_if_missing(genomics_packages)

# Bioconductor packages
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

bioc_packages <- c(
  "VariantAnnotation",  # VCF file handling
  "Biostrings",         # Biological sequences
  "GenomicRanges",      # Genomic intervals
  "SNPlocs.Hsapiens.dbSNP144.GRCh37"
)

# Install Bioconductor packages
for (pkg in bioc_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

# Visualization packages
viz_packages <- c(
  "ggplot2",
  "cowplot",        # Combine ggplot figures
  "ggrepel",        # Better text labels
  "ggridges",       # Ridge plots
  "pheatmap",       # Heatmaps
  "corrplot",       # Correlation plots
  "gridExtra"       # Grid arrangement
)

install_if_missing(viz_packages)

# Load all packages
library(tidyverse)
library(data.table)
library(rrBLUP)
library(qtl)
library(qtl2)
library(ggplot2)
library(pheatmap)
library(corrplot)

message("✓ All packages loaded successfully!")
```

## Project Organization

Create a standardized project structure:

```r
# =============================================================================
# File: create_project_structure.R
# Description: Set up organized folder structure for genomics projects
# =============================================================================

create_genomics_project <- function(project_name, path = ".") {
  
  # Create main directories
  dirs <- c(
    "data/raw",           # Original data (never modify)
    "data/processed",     # Cleaned data
    "data/external",      # Third-party data
    "scripts",            # R scripts
    "output/figures",     # Generated plots
    "output/tables",      # Generated tables
    "output/results",     # Analysis results
    "docs",               # Documentation
    "temp"                # Temporary files
  )
  
  full_paths <- file.path(path, project_name, dirs)
  
  # Create directories
  for (dir in full_paths) {
    if (!dir.exists(dir)) {
      dir.create(dir, recursive = TRUE)
      message(paste("Created:", dir))
    }
  }
  
  # Create .gitignore
  gitignore_content <- c(
    "# R objects",
    "*.RData",
    "*.rds",
    "",
    "# RStudio",
    ".Rhistory",
    ".Rproj.user",
    "",
    "# Temporary files",
    "temp/*",
    "!temp/.gitkeep",
    "",
    "# System files",
    ".DS_Store",
    "Thumbs.db"
  )
  
  writeLines(
    gitignore_content, 
    file.path(path, project_name, ".gitignore")
  )
  
  # Create README
  readme_content <- c(
    paste("#", project_name),
    "",
    "## Description",
    "",
    "Brief description of the project.",
    "",
    "## Directory Structure",
    "",
    "- `data/raw/` - Original data (read-only)",
    "- `data/processed/` - Cleaned data",
    "- `scripts/` - Analysis scripts",
    "- `output/` - Results and figures",
    "",
    "## Setup",
    "",
    "```r",
    "source('scripts/00_setup.R')",
    "```"
  )
  
  writeLines(readme_content, file.path(path, project_name, "README.md"))
  
  message(paste("\n✓ Project", project_name, "created successfully!"))
  message(paste("Location:", file.path(path, project_name)))
}

# Usage:
# create_genomics_project("my_genomics_project")
```

## RStudio Configuration

Optimize RStudio for genomics work:

### Recommended Settings

Go to **Tools → Global Options**:

**Code Tab:**
- ✓ Show line numbers
- ✓ Highlight selected line
- ✓ Auto-indent code
- ✓ Soft-wrap R source files
- Tab width: 2 spaces

**Saving Tab:**
- ✓ Save before sourcing
- Default save format: UTF-8

**Pane Layout Tab:**
- Choose layout that shows:
  - Source editor
  - Console
  - Environment/History
  - Files/Plots/Packages/Help

### Create RStudio Project Template

```r
# =============================================================================
# File: scripts/00_setup.R
# Description: Standard setup for all genomics projects
# =============================================================================

# Set working directory to project root
# (Automatically set when using RStudio Projects)

# Load packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(rrBLUP)
  library(qtl2)
  library(ggplot2)
  library(pheatmap)
})

# Set options
options(
  stringsAsFactors = FALSE,    # Don't convert strings to factors
  scipen = 999,                # Disable scientific notation
  digits = 4,                  # Display 4 decimal places
  width = 120                  # Console width
)

# Set ggplot2 theme
theme_set(theme_minimal(base_size = 12))

# Custom colors for genomics plots
genomics_colors <- c(
  "primary" = "#2b6cb0",
  "secondary" = "#742a2a",
  "success" = "#48bb78",
  "warning" = "#ed8936",
  "error" = "#f56565",
  "neutral" = "#718096"
)

# Helper functions
message_green <- function(...) message(paste0("\033[32m", ..., "\033[0m"))
message_blue <- function(...) message(paste0("\033[34m", ..., "\033[0m"))

message_green("✓ Setup complete!")
```

## Working with Genomic Data

### Example: Reading Different File Types

```r
# =============================================================================
# File: scripts/01_data_io.R
# Description: Input/Output functions for genomic data
# =============================================================================

# Read CSV/TSV files
read_genotype_csv <- function(file_path) {
  read_csv(file_path, show_col_types = FALSE)
}

# Read Excel files
read_phenotype_excel <- function(file_path, sheet = 1) {
  read_excel(file_path, sheet = sheet)
}

# Read PLINK files (using genio)
read_plink_files <- function(bed_file, bim_file, fam_file) {
  if (!require("genio")) install.packages("genio")
  library(genio)
  
  plink_data <- read_plink(
    bed = bed_file,
    bim = bim_file,
    fam = fam_file
  )
  
  return(plink_data)
}

# Read VCF files (using VariantAnnotation)
read_vcf_file <- function(vcf_path) {
  if (!require("VariantAnnotation")) {
    BiocManager::install("VariantAnnotation")
  }
  library(VariantAnnotation)
  
  vcf <- readVcf(vcf_path, "hg19")
  return(vcf)
}

# Write results
write_results_table <- function(data, file_path) {
  write_csv(data, file_path)
  message(paste("✓ Saved:", file_path))
}

# Save R object
save_rds <- function(object, file_path) {
  saveRDS(object, file_path)
  message(paste("✓ Saved RDS:", file_path))
}
```

## Creating Reproducible Reports

### R Markdown Template

```markdown
---
title: "Genomic Analysis Report"
author: "Muhammad Aziz Muslim"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

# Load setup
source("scripts/00_setup.R")
```

## Introduction

Brief description of the analysis.

## Methods

Describe your methods here.

## Results

```{r analysis}
# Your analysis code
```

## Discussion

Interpret your results.
```

## Best Practices

### :::{.callout-important}
**Golden Rules for Reproducible Genomics:**

1. **Never modify raw data** - Always work on copies
2. **Use scripts, not point-and-click** - Document every step
3. **Comment your code** - Future you will thank you
4. **Use version control** - Git + GitHub
5. **Save intermediate results** - Don't re-run everything
6. **Set seeds for randomness** - `set.seed(123)`
:::

### Code Organization Example

```r
# =============================================================================
# File: scripts/02_gwas_analysis.R
# Author: Muhammad Aziz Muslim
# Date: 2024-01-15
# Description: GWAS analysis for plant height
# =============================================================================

# SECTION 1: Load data --------------------------------------------------------
message_blue("Loading data...")
genotype <- readRDS("data/processed/genotype.rds")
phenotype <- read_csv("data/processed/phenotype.csv")

# SECTION 2: Quality control --------------------------------------------------
message_blue("Performing QC...")
# QC code here

# SECTION 3: GWAS analysis ----------------------------------------------------
message_blue("Running GWAS...")
# GWAS code here

# SECTION 4: Visualization ----------------------------------------------------
message_blue("Creating plots...")
# Plotting code here

# SECTION 5: Save results -----------------------------------------------------
message_blue("Saving results...")
write_results_table(results, "output/tables/gwas_results.csv")
```

## Summary

You now have:

- ✅ R and RStudio installed
- ✅ All essential genomics packages
- ✅ Organized project structure
- ✅ Reproducible workflow setup
- ✅ Best practices for coding

## Next Steps

Continue with:

1. [Introduction to Genomics](01-intro-to-genomics.qmd) - If you haven't
2. [GWAS Analysis](03-gwas-analysis.qmd) - Your first analysis

## Troubleshooting

### Package installation fails?

```r
# Try installing from source
install.packages("package_name", type = "source")

# Or use BiocManager for Bioconductor
BiocManager::install("package_name", ask = FALSE)
```

### Memory issues with large datasets?

```r
# Use data.table for efficiency
library(data.table)
large_data <- fread("large_file.csv")

# Or process in chunks
chunk_size <- 10000
```

---

**Citation:** Muslim, M.A. (2024). R Setup for Genomics. *Genomics Tutorials*. https://your-username.github.io/tutorials
